Palmer amaranth (*Amaranthus palmeri*) adaptation to agroecossystems

### 2021-08-04 Manuscript/Manuscript.Rmd
jh

Palmer amaranth (*Amaranthus palmeri* S. Watson) is currently considered one of the most economically damaged weed species to cropping systems in the United States [@ward2013]. The species has showed a remarkable capacity to evolve resistance to herbicides. Palmer amaranth has evolved resistance to eight herbicide sites of action [@heap2021], increasing the weed management complexity [@lindsay2017]. Uncontrolled Palmer amaranth in competition for water, light and nutrients can drastically reduce crop yields [@berger2015]. Palmer amaranth is documented with potential to reduce 91%, 68%, and 54% of corn [@massinga2001], soybean [@klingaman1994], and cotton [@morgan2001] yields, respectively. Thus, Palmer amaranth poses a risk to sustainable agriculture.

Palmer amaranth is a fast growing summer annual forb indigenous to Sonoran Desert [@sauer1957]. The species would eventually emerge as a threat to US agriculture in the 1990s. Palmer amaranth weediness is likely a result of human-assisted selection in combination with species biology. Farm mechanization, conservation agriculture (e.g., no-till), and reliance on herbicides for weed management are the main human-mediated selection of Palmer amaranth into cropping systems. On the other hand, Palmer amaranth is a prolific seed producer with a C4 photosynthetic apparatus [@ward2013]. With a dioecy nature, Palmer amaranth male and female plants are obligate outcrosser species, increasing the chances of exchanging adaptive traits among plants [@oliveira2018]. Also, Palmer amaranth small seed size (e.g, 1 mm) tend to thrive in no-tillage systems [@price2011], and spread across locations through farm equipment [@sauer1972], manure [@hartzler2016], animals [@farmer2017], and plant propagules [@yu2021]. Therefore, Palmer amaranth dispersal capacity make the species one of the most successful cases of weed adaption to cropping systems.

Light and temperature are likely the main environment requirements for Palmer amaranth successful adaptation. Palmer amaranth is reported with an extended germination period [@jha2010]. Germination of Palmer amaranth is triggered by 18 C soil temperature [@keeley1987], and optimal germination and biomass production occur at 35/30 C day and night temperatures [@guo2003]. Water has not shown to limit Palmer amaranth fitness. Under continuous water stress, Palmer amaranth survived and produced at least 14000 seeds plant-1 [@chahal2018]. Also, seeds from Palmer amaranth growing under water stress conditions were heavier, less dormant, and prompt for germination [@matzrafi2021]. The continuous global temperature warming can impact agriculture and promote niches for Palmer amaranth invasion/adaptation into new environments. Currently, it is estimated that the greatest climatic risk of Palmer amaranth establishment are agronomic crops in Australia and Sub-Sahara Africa [@kistner2018]. Temperature is a key factor limiting Palmer amaranth expansion to cooler geographies [@briscoerunquist2019]; however, under future climate change Palmer amaranth is likely to expand northward into Canada and Northern Europe [@kistner2018; @briscoerunquist2019].

Palmer amaranth is already documented in agronomic crops of South America [@larran2017; @kupper2017] and Southern Europe [@milani2021]. In the US, Palmer amaranth is established at crop [@garetson2019] and non-crop land [@bagavathiannan2016] in the warm southern United States but its range is expanding to cool temperatures northward. For example, herbicide resistant Palmer amaranth is widespread in Nebraska  [@oliveira2021a],  Michigan [@kohrt2017], and Connecticut [@aulakh2021]. Successful cases of Palmer amaranth invasion and near to eradication is reported in Minnesota [@yu2021]. No Palmer amaranth actively growing was found in Canada; however, Palmer amaranth seeds was detected in sweet potato slips [@page2021]. Nonetheless, it seems fated the need to manage Palmer amaranth in agronomic crops throughout multiple environments in the near future. Therefore, strategies on Palmer amaranth management should encompass the agroecosystem level but not only attempts to eradicate the weed. Most tactics to manage Palmer amaranth are based on technology fixes [@scott2011], which are short-term (e.g., herbicide and/or tillage) rather than long-term weed management.

The continuous Palmer amaranth dispersal and potential establishment into northern United States warrant investigations on species morphology growing in such environments.  Understanding Palmer amaranth biology and growing strategies under different agroecossystems can enhance our knowledge on species adaptation. It can also aid on designing proactive and ecological tactics to limit the species range expansion, reduce its negative impact, and design resilient and sustainable farming systems [@maclaren2020].  Therefore, the objective of this study was to investigate the flowering pattern, biomass production, and height of Palmer amaranth growing under in corn, soybean and fallow at two timings across five locations in the mid/upper United States Midwest.

@@

Palmer amaranth cumulative flowering from first transplant timing ranged from late June to early September (Figure 2). Palmer amaranth growing in corn resulted in a longer flowering pattern compared to fallow and soybean. Palmer amaranth reached 10% flowering in soybean at doy 180, which was slightly different from fallow (doy 180.9; *P* = 0.01) and corn (doy 181.7; *P* = 0.00). Nonetheless, the 10% cumulative Palmer amaranth flowering in soybean, fallow and corn occurred at the end of June. The 50% Palmer amaranth cumulative flowering occurred in July. For example, Palmer amaranth reached 50% in fallow at doy 193.4, followed by soybean (doy 194.8), corn (doy 206.6). Similar trend was observed at 90% Palmer amaranth cumulative flowering. Palmer amaranth growing in corn reached 90% flowering at doy 252.6 (early September), which was 37.8 and 32.2 days after Palmer amaranth 90% flowering in fallow and soybean, respectively.

@@

Palmer amaranth cumulative flowering from second transplanting timing ranged from mid July to mid September (Figure 2). All cumulative flowering comparisons among agroecossystems were significant (*P* < 0.00) Palmer amaranth growing in fallow resulted in earlier flowering time compared to soybean and corn. Palmer amaranth growing in fallow reached 10%, 50%, and 90% flowering time at day 203.8, 214.4, and 232.2, respectively. Palmer amaranth growing in soybean reached 10% flowering at doy 210.9, which was 6 days prior to corn (*P*-value = 0.00). Similar trend was observed at 50% flowering, whereas Palmer amaranth reached 50% flowering in corn (doy 233.0) 4 days after soybeans (doy 228.9; *P* = 0.00). The 90% Palmer amaranth cumulative flowering occurred at same day in corn (260.9) and soybean (260.5; *P* = 0.66).

@@

Palmer amaranth accumulated more biomass when growing in fallow compared to Palmer amaranth growing in soybean and corn. At first transplanting time, Palmer amaranth biomass was 78.3 g plant-1, 26.4 g plant-1 and 14.9 g plant-1 in fallow, soybean and corn, respectively. Palmer amaranth produce less biomass at second transplanting time. For example, Palmer amaranth growing in fallow resulted in 49.2 g plant-1, 3.4 g plant-1 in soybean and 1.4 g plant when growing in corn.

@@

>>>>

Palmer amaranth (*Amaranthus palmeri* S. Watson) is currently considered one of the most economically damaged weed species to cropping systems in the United States. Unmanaged Palmer amaranth in competition for water, light and nutrients can drastically impact on crop yields [@berger2015]. For example, Palmer amaranth is documented with potential to reduce 91%, 68%, and 54% of corn [@massinga2001], soybean [@klingaman1994], and cotton [@morgan2001] yields, respectively. Moreover, Palmer amaranth has showed a remarkable capacity to evolve resistance to herbicides. To date, Palmer amaranth has evolved resistance to eight herbicide sites of action [@heap2021], increasing the weed management complexity [@lindsay2017]. Thus, Palmer amaranth poses an economical risk to sustainable agriculture.

Palmer amaranth is a fast growing summer annual forb indigenous to Sonoran Desert [@sauer1957]. The species would eventually emerge as a threat to US agriculture in the 1990s. Palmer amaranth weediness is likely a result of human-assisted selection in combination with species biology. Farm mechanization, conservation agriculture (e.g., no-till), and reliance on herbicides for weed management are the main human-mediated selection of Palmer amaranth into cropping systems [@ward2013]. On the other hand, Palmer amaranth is a prolific seed producer with a C4 photosynthetic apparatus [@wang1992]. With a dioecy nature, Palmer amaranth male and female plants are obligate outcrosser species, increasing the chances of exchanging adaptive traits among plants [@oliveira2018]. Also, Palmer amaranth small seed size (e.g, 1 mm) tend to thrive in no-tillage systems [@price2011], and spread across locations through farm equipment [@sauer1972], manure [@hartzler2016], animals [@farmer2017], and plant propagules [@yu2021]. The dispersal capacity of Palmer amaranth make the species one of the most successful cases of weed adaption to cropping systems.

Light and temperature are likely the main environment requirements for Palmer amaranth successful grow and development. Palmer amaranth is reported with an extended germination period [@jha2010]. Germination of Palmer amaranth is triggered by 18 C soil temperature [@keeley1987], and optimal germination and biomass production occur at 35/30 C day and night temperatures [@guo2003]. Also, Palmer amaranth germination are human-mediated by tillage timings and preemergence-applied herbicides [@chahal2021]. Water has not shown to limit Palmer amaranth fitness. Under continuous water stress, Palmer amaranth survived and produced at least 14000 seeds plant-1 [@chahal2018]. Also, seeds from Palmer amaranth growing with limited water conditions were heavier, less dormant, and prompt for germination [@matzrafi2021]. The continuous global temperature warming can impact agriculture and promote niches for Palmer amaranth invasion/adaptation into new environments. Currently, it is estimated that the greatest climatic risk of Palmer amaranth establishment are agronomic crops in Australia and Sub-Sahara Africa [@kistner2018]. Temperature is a key factor limiting Palmer amaranth expansion to cooler geographies [@briscoerunquist2019]; however, under future climate change Palmer amaranth is likely to expand northward into Canada and Northern Europe [@kistner2018; @briscoerunquist2019].

Palmer amaranth is already found in agronomic crops of South America [@larran2017; @kupper2017] and Southern Europe [@milani2021]. In the US, Palmer amaranth is well established at cotton fields [@garetson2019] and non-crop areas [@bagavathiannan2016] in the southern United States but its range is expanding northward. For example, herbicide resistant Palmer amaranth is widespread in Nebraska  [@oliveira2021a],  Michigan [@kohrt2017], Connecticut [@aulakh2021], and potential damaged to soybean fields in Illinois is reported [@davis2015]. Successful cases of Palmer amaranth invasion and near to eradication is documented in Minnesota [@yu2021]. No Palmer amaranth actively growing was found in Canada; however, Palmer amaranth seeds was detected in sweet potato slips [@page2021]. Nonetheless, it seems fated the need to manage Palmer amaranth in agronomic crops throughout multiple environments in the near future. Strategies on Palmer amaranth management should encompass the agroecosystem level but not only attempts to eradicate the weed. Most tactics to manage Palmer amaranth are based on technology fixes [@scott2011], which are short-term (e.g., herbicide and/or tillage) rather than long-term weed management. Palmer amaranth management should be built on minimizing the species ability to adapt, grow and develop into agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to have higher impact on cotton yields than late established plants [@macrae2013]. In the northern states, Palmer amaranth impact on the agroecossystem is recent. Studies investigating Palmer amaranth in those locations is limited due to the plant classification as noxious weed [@yu2021]. Nonetheless, the continuous Palmer amaranth dispersal and potential establishment across northern United States is concerning and warrant investigations on species morphology in such environments. Understanding Palmer amaranth biology and growing strategies under different agroecossystems can enhance our knowledge on species adaptation. It can also aid on designing proactive and ecological tactics to limit the species range expansion, reduce its negative impact, and design resilient and sustainable farming systems [@maclaren2020]. Therefore, the objective of this study was to investigate the flowering pattern, biomass production, and height of Palmer amaranth growing under corn, soybean and fallow at two timings across five locations in the United States Midwest.

@@

Palmer amaranth growing in corn resulted in a longer flowering pattern compared to fallow and soybean at first cohort (Figure 2A). The 10% cumulative Palmer amaranth flowering in soybean, fallow and corn occurred at the end of June. Palmer amaranth reached 10% flowering in soybean at doy 180, which was slightly different from fallow (doy 180.9; *P* = 0.01) and corn (doy 181.7; *P* = 0.00). The 50% Palmer amaranth cumulative flowering occurred in July. For example, Palmer amaranth reached 50% in fallow at doy 193.4, followed by soybean (doy 194.8), corn (doy 206.6). Similar trend was observed at 90% Palmer amaranth cumulative flowering. Palmer amaranth growing in corn reached 90% flowering at doy 252.6 (early September), which was 37.8 and 32.2 days after Palmer amaranth 90% flowering in fallow and soybean, respectively.

@@

Palmer amaranth cumulative flowering at second cohort ranged from mid July to mid September (Figure 2). All cumulative flowering comparisons among agroecossystems were significant different (*P* < 0.00) Palmer amaranth growing in fallow resulted in earlier flowering time compared to soybean and corn. Palmer amaranth growing in fallow reached 10%, 50%, and 90% flowering time at day 203.8, 214.4, and 232.2, respectively. Palmer amaranth growing in soybean reached 10% flowering at doy 210.9, which was 6 days prior to corn (*P*-value = 0.00). Similar trend was observed at 50% flowering, whereas Palmer amaranth reached 50% flowering in corn (doy 233.0) 4 days after soybeans (doy 228.9; *P* = 0.00). The 90% Palmer amaranth cumulative flowering occurred at same day in corn (260.9) and soybean (260.5; *P* = 0.66).

@@

Palmer amaranth accumulated more biomass when growing in fallow compared to Palmer amaranth growing in soybean and corn (Figure 3A). At first cohort time, Palmer amaranth biomass was 78.3 g plant-1, 26.4 g plant-1 and 14.9 g plant-1 in fallow, soybean and corn, respectively. Moreover, Palmer amaranth produce less biomass at cohort time. Palmer amaranth growing in fallow resulted in 49.2 g plant-1, 3.4 g plant-1 in soybean and 1.4 g plant when growing in corn.

The Palmer amaranth achieve 92.3 cm tall when growing at first cohort timing in corn (Figure 3B).

@@

Our study showed that Palmer amaranth flowering time, biomass and height varied within agroecossystems and cohort timing. In general, Palmer amaranth produced more biomass and taller plants when growing at first cohort rather than second cohort. At first cohort, resources (e.g., soil nutrients) and conditions (e.g., light) were more timely available for both species, crop and weed. High biomass and taller Palmer amaranth plants is likely a weed strategy to compete for light in between crop rows in absence of canopy. In such condition, Palmer amaranth showed an extraordinary plasticity to adapt upon the agroecosystem. This is evident when comparing Palmer amaranth shape, and its extended flowering pattern when growing into corn compared to soybean. In corn, Palmer amaranth was taller and thinner when growing in corn. The Palmer amaranth competition strategy was likely to mimic the crop grow and development (Figure 4). These results are concerning as breeding more competitive crop varieties is likely to select more competitive Palmer amaranth biotypes [@bravo2017].

Palmer amaranth grow and development at second cohort was limited due to the crop competitive ability at advanced development stages. Palmer amaranth was transplanted when corn canopy was nearly closed, which reduced Palmer amaranth competitiveness. As a result, Palmer amaranth height and biomass was lower compared to its first cohort. Under crop canopy, Palmer amaranth flowering pattern was near to similar into corn and soybean. Palmer amaranth growing without crop competition produced highest amounts of biomass and lowest extended flowering pattern. The Palmer amaranth strategy in bareground was to invest biomass in growing plant width and height. Nonetheless, Palmer amaranth produced 31% less biomass in second cohort compared to first cohort timing. In a bareground study, early emerged Palmer amaranth without competition was 50% taller than late emerged plants [@webster2015]. These results suggests that crop competition is not the only factor limiting late Palmer amaranth establishment. The limited growth of Palmer amaranth at second cohort is likely a reduced plant response to thermal units (e.g, growing degree days). It is hypothesize that reduced day length contributed to smaller plants at second cohort as well as shorter flowering period. A study in North Carolina and Illinois predicted that less than 10% Palmer amaranth seedlings emergence occurred after June [@piskackova2021]. In addition, Palmer amaranth negative impact on soybean [@korres2020] and cotton [@webster2015] yields was higher when plants were established near to crop planting.

Seed production was not evaluated due to Palmer amaranth harvest at flowering time. Nonetheless, it is well documented a strong positive correlation between Palmer amaranth biomass and seed production [@schwartz2016; @spaunhorst2018]. I our study, Palmer amaranth growing at first cohort accumulated an overall 50% more biomass when compared to second cohort. Therefore, Palmer amaranth plants growing in the second cohort is likely to produce less seeds regardless the agroecossystem. Our observation is consistent with the findings that first Palmer amaranth cohort produced 50% more seeds per plant than Palmer amaranth plants established six weeks later in bareground [@webster2015]. Nonetheless, seed production at second cohort is likely to replenish the soil seedbank. Seed production and deposition in the seedbank is a key factor for species perpetuation [@menges1987]. Palmer amaranth can produce hundred thousands seeds per plant [@schwartz2016; @keeley1987], and stay viable buried in the seedbank for at least 36 months [@sosnoskie2013].

Our study demonstrated the short-term Palmer amaranth plasticity to grow and develop into cropping-systems. Palmer amaranth management priority should focus on minimizing Palmer amaranth dispersal into new geographies. Early-season management programs would have a large negative effect in Palmer amaranth growth and development. Tactics that promote crop advantage against Palmer amaranth, including early crop planting, crop rotation, plant width, preemergence applied herbicide [@sanctis2021], and crop residue (e.g. cover crops) would minimize the negative impact of Palmer amaranth in agroecossystems.
